Relationships in the extremes and their influence on competition and coexistence: Supporting information

### 2021-07-06 paper/SupMat.Rnw
further progress in writing sup mat sections associated with Methods

\section{Scaling factors}\label{SIsect:scaling_factors}
%***DAN: I am not completely sure about anything in this section and need to get Ellner to look at it. I 
%did give it a reasonable shot though.

@@

\cite{Chesson_1994} and \cite{Ellner_2016} use $q_{ij}=1$ for this same model. Why the discrepancy? 
DAN: You need to explain this discrepancy here after talking to Ellner.
%***DAN: Add some text explaining the discrepancy

@@

As in the main text, we can now define the tertiary decomposition
\begin{align}
\olep_i^{(EC)} &= \E[r_i(E_i,C_{i \bs i})]-\E[r_i(E_i^\shp,C_{i \bs i}^\shp)] \\
&= \left[ \E[r_i(E_i,C_{i \bs i})]-\E[r_i(E_i^\nat,C_{i \bs i}^\nat)] \right]+\left[ \E[r_i(E_i^\nat,C_{i \bs i}^\nat)]-\E[r_i(E_i^\shp,C_{i \bs i}^\shp)] \right] \\
&= \olep_i^{[E \nat C]}+\olep_i^{[EC]}.
\end{align}
The brackets in the superscripts of $\olep_i^{[EC]}$ and $\olep_i^{[E \nat C]}$ indicate that we are talking about a term in the
tertiary decomposition, i.e., in the decomposition of $\olep_i^{(EC)}$. 
The term $\olep_i^{[E \nat C]}$ quantifies the influence of asymmetric tail associations of $E_i$ and $C_{i \bs i}$ on the invasion growth rate,
and the the term $\olep_i^{[EC]}$ corresponds to $EC$ correlations \emph{per se}, after the removal of asymmetries of tail association.
The symbol $\nat$ denotes the removal of asymmetries of tail association. We therefore have 
\begin{equation}
\olr_{i \bs i} = \epsilon_i^0 + \olep_i^E + \olep_i^C + \olep_i^{(E\shp C)}+\olep_i^{[EC]}+\olep_i^{[E \nat C]},\label{eq:decomp_overall_IGR}
\end{equation}
our overall decompsition of $r_{i \bs i}$.
The same derivation also applies to provide 
\begin{equation}
\olr_{j \bs i} = \epsilon_j^0 + \olep_j^E + \olep_j^C + \olep_j^{(E\shp C)}+\olep_j^{[EC]}+\olep_j^{[E \nat C]},
\end{equation}
an analogous decomposition of $\olr_{j \bs i}$.

Combining the decompositions of $\olr_{i \bs i}$ and $\olr_{j \bs i}$ and assuming without loss of generality that $\mu_i \leq \mu_j$ so that
species $i$ is the weaker competitor and $\olr_{i \bs i}-q_{ij}\olr_{j \bs i}$ can be regarded as a coexistence metric, we have
\begin{equation}
\olr_{i \bs i}-q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[EC]}+\Delta_i^{[E\nat C]},
\end{equation}
where $\Delta_i^0=\epsilon_i^0- q_{ij} \epsilon_j^0$;
$\Delta_i^E=\olep_i^E -q_{ij}\olep_j^E$ is the contribution to coexistence of environmental variation;
$\Delta_i^C=\olep_i^C -q_{ij}\olep_j^C$ is the contribution to coexistence of variation in competitive pressure;
$\Delta_i^{(E\shp C)}=\olep_i^{(E\shp C)} - q_{ij}\olep_j^{(E\shp C)}$ is the contribution to coexistence of 
variation \emph{per se} in both environment and competition, not including the effects of covariation of these quantities;
$\Delta_i^{[EC]}=\olep_i^{[EC]} - q_{ij} \olep_j^{[EC]}$ is the contribution to coexistence of $EC$ covariation \emph{per se},
not including the effects of asymmetric tail associations between environment and competition;
and $\Delta_i^{[E\nat C]}=\olep_i^{[E \nat C]}-q_{ij} \olep_j^{[E \nat C]}$ is the contribution to coexistence of 
asymmetric tail assocations. The subscript $i$ on all these newly defined quantities $\Delta_i$ refers to species $i$.

@@

When needed, noise with symmetric tail association was generated by taking a very large sample (100000) of left-tail associated noise generated as 
described above, and then calculating the Pearson correaltion coefficient, $\rho$, of the sample.
Symmetric noise $\left(\tilde b_1^{(i)},\tilde b_2^{(i)}\right)$ for $i=1,...,M$ was then generated 
from the bivariate normal distribution $N\left(\vec{0}, \Sigma \right)$, where $\Sigma = \begin{pmatrix} 1&\rho \\ \rho&1\end{pmatrix}$, 
and then we set $b_1^{(i)} = \sigma \tilde b_1^{(i)} + \mu_1$ and $b_2^{(i)} = \sigma \tilde b_2^{(i)}+ \mu_2$.

@@

For the lottery model, we can straightforwardly show from the formulas of SI section \ref{SIsect:other_mechanisms} that:

@@

&= \E[\ln(1-\delta+\delta\exp(b_i-b_j))]

@@

@@

\E[\ln(1-\delta+\delta\exp(\sigma (\tilde{b}_i^\shp - \tilde{b}_j^\shp)+\mu_i-\mu_j))].

@@

%***DAN: Next think it all through for j\i and determine what needs to be said for that case before combining the two cases 
%(and also scaling factors) to get the delta terms in extended MTC decomp
%Once you do those things you can observe that all these terms depend only on mui-muj and not independently on mui or muj.

>>>>

\section{Scaling factors}\label{SIsect:scaling_factors}

@@

%***DAN: Change the below paragraph if needed. Some reasons it might be needed: Steve joins as a coauthor;
%Steve reads this and thinks it makes sense to change it because something is factually or presentationally 
%sub-optimal; or Dan or Jasmin actually develops a more rigorous understanding of scaling factors and based
%on that thinks something here is sub-optimal. Anyway, the "do both" approach seems like it can't go wrong,
%especially if Steve's prediction is correct that it won't make much difference.
%***DAN: You probably need to spend a bit more time with the cited papers to better understand this. 
We also use $q_{ij}=1$ as an alternative value, based on communication with Steve Ellner, for the following reasons.
First, as discussed in the supporting information of \cite{Ellner_2016}, and as further elaborated by 
\cite{Barabas_2018}, it takes very special circumstances for the scaling factors to exist and be unique, 
especially with size-structured models. Second, in empirically parameterized models, scaling factor values are 
often not robust to arbitrary modelling choices. 
Although the lottery model is not size structured and we do not empirically parameterize it, the above shortcomings of 
scaling factors, generally, suggest we should find alternatives to their use. The alternative 
suggested by Ellner for empirical applications is to do a series of pairwise invader-resident comparisons, each with an 
equal weighting of invader and resident $r$ values; for our context this amounts to setting scaling factors to $1$. 
We consider both approaches because the above-mentioned shortcomings do not apply to our case, but it seems 
possible that future applications will often set scaling factors to $1$, and we want our ideas to also be considered 
by researchers making that choice.

@@

As in the main text, we can now define the tertiary decomposition
\begin{align}
\olep_i^{(EC)} &= \E[r_i(E_i,C_{i \bs i})]-\E[r_i(E_i^\shp,C_{i \bs i}^\shp)] \\
&= \left[ \E[r_i(E_i,C_{i \bs i})]-\E[r_i(E_i^\nat,C_{i \bs i}^\nat)] \right]+\left[ \E[r_i(E_i^\nat,C_{i \bs i}^\nat)]-\E[r_i(E_i^\shp,C_{i \bs i}^\shp)] \right] \\
&= \olep_i^{[E \nat C]}+\olep_i^{[EC]}.
\end{align}
The brackets in the superscripts of $\olep_i^{[EC]}$ and $\olep_i^{[E \nat C]}$ indicate that we are talking about a term in the
tertiary decomposition, i.e., in the decomposition of $\olep_i^{(EC)}$. 
The term $\olep_i^{[E \nat C]}$ quantifies the influence of asymmetric tail associations of $E_i$ and $C_{i \bs i}$ on the invasion growth rate,
and the the term $\olep_i^{[EC]}$ corresponds to $EC$ correlations \emph{per se}, after the removal of asymmetries of tail association.
The symbol $\nat$ denotes the removal of asymmetries of tail association. We therefore have 
\begin{equation}
\olr_{i \bs i} = \epsilon_i^0 + \olep_i^E + \olep_i^C + \olep_i^{(E\shp C)}+\olep_i^{[EC]}+\olep_i^{[E \nat C]},\label{eq:decomp_overall_IGR}
\end{equation}
our overall decompsition of $r_{i \bs i}$.
The same derivation also applies to provide 
\begin{equation}
\olr_{j \bs i} = \epsilon_j^0 + \olep_j^E + \olep_j^C + \olep_j^{(E\shp C)}+\olep_j^{[EC]}+\olep_j^{[E \nat C]},\label{eq:decomp_overall_rjbsi}
\end{equation}
an analogous decomposition of $\olr_{j \bs i}$.

Combining the decompositions of $\olr_{i \bs i}$ and $\olr_{j \bs i}$ and assuming without loss of generality that $\mu_i \leq \mu_j$ so that
species $i$ is the weaker competitor and $\olr_{i \bs i}-q_{ij}\olr_{j \bs i}$ can be regarded as a coexistence metric, we have
\begin{equation}
\olr_{i \bs i}-q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[EC]}+\Delta_i^{[E\nat C]},\label{eq:decomp_overall_delta}
\end{equation}
where $\Delta_i^0=\epsilon_i^0- q_{ij} \epsilon_j^0$;
$\Delta_i^E=\olep_i^E -q_{ij}\olep_j^E$ is the contribution to coexistence of environmental variation;
$\Delta_i^C=\olep_i^C -q_{ij}\olep_j^C$ is the contribution to coexistence of variation in competitive pressure;
$\Delta_i^{(E\shp C)}=\olep_i^{(E\shp C)} - q_{ij}\olep_j^{(E\shp C)}$ is the contribution to coexistence of 
variation \emph{per se} in both environment and competition, not including the effects of covariation of these quantities;
$\Delta_i^{[EC]}=\olep_i^{[EC]} - q_{ij} \olep_j^{[EC]}$ is the contribution to coexistence of $EC$ covariation \emph{per se},
not including the effects of asymmetric tail associations between environment and competition;
and $\Delta_i^{[E\nat C]}=\olep_i^{[E \nat C]}-q_{ij} \olep_j^{[E \nat C]}$ is the contribution to coexistence of 
asymmetric tail assocations. The subscript $i$ on all these newly defined quantities $\Delta_i$ refers to species $i$.

@@

When needed, noise with symmetric tail association was generated by taking a very large sample (100000) 
%***DAN: Dan, later, check whether 100000 is what was actually used. I'm thinking we could probably do a million.
of left-tail associated noise generated as 
described above, and then calculating the Pearson correaltion coefficient, $\rho$, of the sample.
Symmetric noise $\left(\tilde b_1^{(i)},\tilde b_2^{(i)}\right)$ for $i=1,...,M$ was then generated 
from the bivariate normal distribution $N\left(\vec{0}, \Sigma \right)$, where $\Sigma = \begin{pmatrix} 1&\rho \\ \rho&1\end{pmatrix}$, 
and then we set $b_1^{(i)} = \sigma \tilde b_1^{(i)} + \mu_1$ and $b_2^{(i)} = \sigma \tilde b_2^{(i)}+ \mu_2$.

@@

We first compute all terms of the decomposition (\ref{eq:decomp_overall_IGR}) of $\olr_{i \bs i}$ for the lottery model. 
One can straightforwardly show from the formulas of SI section \ref{SIsect:other_mechanisms} that

@@

&= \E[\ln(1-\delta+\delta\exp(b_i-b_j))]\label{eq:base_IGR}

@@

and

@@

\E[\ln(1-\delta+\delta\exp(\sigma (\tilde{b}_i^\shp - \tilde{b}_j^\shp)+\mu_i-\mu_j))].\label{eq:decomp:contrib_EC}

@@

We next decompose $\olr_{j \bs i}$ for the lottery model. We have 
E_j = B_j = \exp(b_j)
C_j=\frac{B_1 N_1 + B_2 N_2}{\delta N}
C_{j \bs i} = B_j/\delta = \exp(b_j)/\delta
r_j=\ln(1-\delta+E_j/C_j)
\olr_{j \bs i} &= \E[\ln(1-\delta+E_j/C_{j \bs i})] \\
&= \E[\ln(1-\delta+\delta B_j/B_j)] \\
&= 0
\overline{E}_j = \exp(\mu_j+\sigma^2/2)
\overline{C}_{j \bs i} = \exp(\mu_j + \sigma^2/2)/\delta
\epsilon_j^0 &= r_j(\overline{E}_j,\overline{C}_{j \bs i}) \\
&= \ln \left( 1-\delta+\delta \frac{\exp(\mu_j + \sigma^2/2)}{\exp(\mu_j + \sigma^2/2)} \right) \\
&= 0 \label{eq:decomp:baseline_j}
and
\olep_j^E &= \E[r_j(E_j,\overline{C}_{j \bs i})]-\epsilon_j^0 \\
&= \E[\ln(1-\delta+\delta B_j/\overline{B}_j)]-\epsilon_j^0 \\
&= \E \left[ \ln \left(1-\delta+\delta \frac{\exp(b_j)}{\exp(\mu_j+\sigma^2/2)} \right) \right]-\epsilon_j^0 \\
&= \E \left[ \ln \left(1-\delta+\delta\frac{\exp(\sigma \tilde{b}_j + \mu_j)}{\exp(\mu_j+\sigma^2/2)} \right) \right]-\epsilon_j^0 \\
&= \E[\ln(1-\delta+\delta \exp(\sigma \tilde{b}_j - \sigma^2/2))]-\epsilon_j^0, \label{eq:decomp:contrib_of_var_E_j}
\olep_j^C &= \E[r_j(\overline{E}_j,C_{j \bs i})]-\epsilon_j^0 \\
&= \E \left[ \ln \left(1-\delta+\delta \frac{\overline{B}_j}{B_j} \right) \right]-\epsilon_j^0 \\
&= \E \left[ \ln \left(1-\delta+\delta \frac{\exp(\mu_j+\sigma^2/2)}{\exp(b_j)} \right) \right]-\epsilon_j^0 \\
&= \E \left[ \ln \left(1-\delta+\delta \frac{\exp(\mu_j+\sigma^2/2)}{\exp(\sigma \tilde{b}_j +\mu_j)} \right) \right]-\epsilon_j^0 \\
&= \E[\ln(1-\delta+\delta \exp(-\sigma \tilde{b}_j+\sigma^2/2))]-\epsilon_j^0,\label{eq:decomp:contrib_of_var_C_j}
where again $\tilde{b}_j$ is standard normally distributed (see SI section \ref{sect:noise}). Continuing,
\olep_j^{EC} &= \E[r_j(E_j,C_{j \bs i})]-\olep_j^E-\olep_j^C-\epsilon_j^0 \\
&= \E[\ln(1-\delta+\delta B_j/B_j)]-\olep_j^E-\olep_j^C-\epsilon_j^0 \\
&= -\olep_j^E-\olep_j^C-\epsilon_j^0.\label{eq:decomp:j:EC}
Continuing,
\olep_j^{(E\shp C)} &= \E[r_j(E_j^\shp , C_{j \bs i}^\shp )]-\olep_j^E-\olep_j^C-\epsilon_j^0 \\
&= \E \left[ \ln \left(1-\delta+\delta \frac{B_j^\shp}{B_j} \right) \right]-\olep_j^E-\olep_j^C-\epsilon_j^0 \\
&= \E \left[ \ln \left(1-\delta+\delta \frac{\exp(b_j^\shp)}{\exp(b_j)} \right) \right]-\olep_j^E-\olep_j^C-\epsilon_j^0 \\
&= \E \left[ \ln \left(1-\delta+\delta \frac{\exp(\sigma \tilde{b}_j^\shp + \mu_j)}{\exp(\sigma \tilde{b}_j + \mu_j)} \right) \right]-\olep_j^E-\olep_j^C-\epsilon_j^0 \\
&= \E[\ln(1-\delta+\delta \exp(\sigma(\tilde{b}_j^\shp - \tilde{b}_j)))]-\olep_j^E-\olep_j^C-\epsilon_j^0, \label{eq:decomp:j:EshpC}
where $(E_j^\shp , C_{j \bs i}^\shp)$ has the same marginals as $(E_j,C_{j \bs i})$ but has independent components;
$B_j^\shp$ is distributed in the same way as $B_j$ but is independent of it; $b_j^\shp$ is distributed in the same
way as $b_j$ but is independent of it; and $\tilde{b}_j$ and $\tilde{b}_j^\shp$ are standard normally distributed and 
independent of each other. Continuing,
\olep_j^{(EC)} &= \E[r_j(E_j,C_{j \bs i})]-\E[r_j(E_j^\shp , C_{j \bs i}^\shp)] \\
&= -\E[r_j(E_j^\shp , C_{j \bs i}^\shp)] \\
&= -\E[\ln(1-\delta+\delta \exp(\sigma (\tilde{b}_j^\shp - \tilde{b}_j)))],
these results obtained from (\ref{eq:decomp:j:EC}) and (\ref{eq:decomp:j:EshpC}). Again, $\tilde{b}_j$ and 
$\tilde{b}_j^\shp$ are standard normally distributed and independent. Continuing,
\olep_j^{[E \nat C]} &= \E[r_j(E_j,C_{j \bs i})] - \E[r_j(E_j^\nat , C_{j \bs i}^\nat)] \\
&= \olr_{j \bs i} - \olr_{j \bs i}^\nat \\
&= 0
because we showed in the Theory section of the main text that $\olr_{j \bs i}^\nat = \olr_{j \bs i}$ due to the fact that
$E_j$ and $C_{j \bs i}$ are perfectly linearly related to each other. Continuing,
\olep_j^{[EC]} &= \olep_j^{(EC)} - \olep_j^{[E \nat C]} \\
&= - \E[\ln(1-\delta+\delta \exp(\sigma(\tilde{b}_j^\shp - \tilde{b}_j)))],
where again $\tilde{b}_j$ and $\tilde{b}_j^\shp$ are standard normally distributed and independent of each other.
This completes the calculation of all the terms of the decomposition of $\olr_{j \bs i}$ for the lottery model.

Using the above formulas, we observe that all the terms of the decomposition of $\olr_{i \bs i}$ depend on 
$\delta$, $\sigma$, and $\mu_i - \mu_j$, but not separately on $\mu_i$ or $\mu_j$. The terms of the decomposition
of $\olr_{j \bs i}$ depend on $\delta$ and $\sigma$, but do not depend on $\mu_i$ or $\mu_j$ at all.
As discussed in the main text and SI section \ref{SIsect:scaling_factors}, we consider two alternative
values for the scaling factors, $q_{ij}$, namely $q_{ij}=1$ and 
\begin{align}
q_{ij} &= \frac{\exp(\mu_i)}{(1-\delta) \exp(\mu_j)+\delta \exp(\mu_i)} \\
&= \frac{\exp(-\mu_i)\exp(\mu_i)}{\exp(-\mu_i)[(1-\delta) \exp(\mu_j)+\delta \exp(\mu_i)]} \\
&= \frac{1}{(1-\delta)\exp(-(\mu_i-\mu_j))+\delta}.
\end{align}
These also either do not depend on $\mu_i$ or $\mu_j$ at all or they depend on the difference 
$\mu_i-\mu_j$. Thus each of the terms of the decomposition (\ref{MT-eq:full_decomp}) from the main text
depends only on $\mu_i-\mu_j$ and not separately on $\mu_i$ or $\mu_j$, as stated in the Methods section of the main text.

We now want to show that, for the lottery model, 
all elements of our decompositions (\ref{eq:decomp_overall_IGR}) and (\ref{eq:decomp_overall_rjbsi}), as well
as $q_{ij}$, and therefore all elements of our decomposition (\ref{eq:decomp_overall_delta}), are the same 
for our left- and right-tail associated cases, as was stated in Methods in the main text. Using the notation 
$(E_i^{(l)},C_{i \bs i}^{(l)})=(B_i^{(l)},B_j^{(l)}/\delta)=(\exp(b_i^{(l)}),\exp(b_j^{(l)})/\delta)$ and 
$(E_i^{(r)},C_{i \bs i}^{(r)})=(B_i^{(r)},B_j^{(r)}/\delta)=(\exp(b_i^{(r)}),\exp(b_j^{(r)})/\delta)$
for the left- and right-tail associated cases, respectively, we note that the marginal distributions of
$(b_i^{(l)},b_j^{(l)})$ are the same as those of $(b_i^{(r)},b_j^{(r)})$, and likewise the marginals of 
$(B_i^{(l)},B_j^{(l)}/\delta)$ are the same as those of $(B_i^{(r)},B_j^{(r)}/\delta)$.
So any component of (\ref{eq:decomp_overall_IGR}), or $q_{ij}$, will not depend on the choice
of our left- or right-tail associated noise cases as long as it depends only on marginal distributions.
But it is easy to see that $q_{ij}$ depends only on marginals, since $q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$  
depends only on $\mu_i$, $\mu_j$ and $\delta$, and these are parameters of the marginal distributions of 
$(b_i,b_j)$. The alternative value $q_{ij}=1$ is constant, so also takes the same value for both our
left- and right-tail association noise cases. It is also easy to see from the definitions of $\epsilon_i^0$, 
$\olep_i^E$, $\olep_i^C$, and $\olep_i^{(E \shp C)}$ that these quantities only depend on marginals.

Whereas $\olep_i^{[EC]} = \E[r_i(E_i^\nat , C_{i \bs i}^\nat )]-\E[r_i(E_i^\shp ,C_{i \bs i}^\shp)]$ depends on
more than just marginals, this quantity is still independent of the choice between our left- and right-tail-association 
cases because the bivariate random variables $(E_i^\nat , C_{i \bs i}^\nat )$ and 
$(E_i^\shp ,C_{i \bs i}^\shp)$ have had their tail association removed, and the only differences 
between $(E_i^{(l)},C_{i \bs i}^{(l)})$ and $(E_i^{(r)},C_{i \bs i}^{(r)})$ are in their tail association
properties (i.e., these two bivariate distributions have the same marginals and covariances).
In other words, $(E_i^\nat , C_{i \bs i}^\nat )$ is distributed in the same way for the left- and right-tail-association cases,
as is $(E_i^\shp ,C_{i \bs i}^\shp)$.

We now examine $\olep_i^{[E \nat C ]}$, which, by (\ref{eq:decomp:contrib_ATA_1}), equals 
\begin{equation}
\olep_i^{[E \nat C]} = \E[\ln(1-\delta+\delta\exp(b_i-b_j))]-\E[\ln(1-\delta+\delta\exp(\ddot{b}_i-\ddot{b}_j))].\label{eq:rep1}
\end{equation}
But $(\ddot{b}_i,\ddot{b}_j)$ is a bivariate normal random variable constructed to have the same marginals and 
covariance as $(b_i,b_j)$, and the marginals and covariance of $(b_i^{(l)},b_j^{(l)})$ are the same as those of
$(b_i^{(r)},b_j^{(r)})$, so $(\ddot{b}_i^{(l)},\ddot{b}_j^{(l)})$ is distributed in the same way as 
$(\ddot{b}_i^{(r)},\ddot{b}_j^{(r)})$. So the second term on the right of (\ref{eq:rep1}) is independent
of the choice between our left- and right-tail-associated noise cases. Now examining 
$\E[\ln(1-\delta+\delta\exp(b_i-b_j))]$, we define $\beta_1=-b_2+\mu_1+\mu_2$ and $\beta_2=-b_1+\mu_1+\mu_2$,
and we note that $(\beta_1^{(l)},\beta_2^{(l)})$ is distributed in the same way as 
$(b_1^{(r)},b_2^{(r)})$; likewise, $(\beta_1^{(r)},\beta_2^{(r)})$ is distributed in the same way as 
$(b_1^{(l)},b_2^{(l)})$. But then we have 
\begin{align}
\E[\ln(1-\delta+\delta\exp(b_i^{(l)}-b_j^{(l)}))] &= \E[\ln(1-\delta+\delta\exp(-\beta_j^{(l)}+\mu_1+\mu_2-(-\beta_i^{(l)}+\mu_1+\mu_2)))] \\
&= \E[\ln(1-\delta+\delta\exp(\beta_i^{(l)}-\beta_j^{(l)}))] \\
&= \E[\ln(1-\delta+\delta\exp(b_i^{(r)}-b_j^{(r)}))], 
\end{align}
which completes the proof that $\olep_i^{[E \nat C]}$ is independent of left- versus right-tail associated noise.
For the terms of the decomposition (\ref{eq:decomp_overall_rjbsi}), it suffices to look at the equations for those terms, 
above, in this section, and note that all those terms depend only on marginals.

\section{Efficient computation}\label{sect:efficient_computation}

We here elaborate how the terms in the decompositions (\ref{eq:decomp_overall_IGR}), (\ref{eq:decomp_overall_rjbsi})
and (\ref{eq:decomp_overall_delta}) are computed rapidly, for the lottery model.

First, for a large integer $M$ (we used $M=$),
%***DAN: fill in the value, 1000000 if you can swing it
carry out the following steps once.
\begin{enumerate}
\item Generate $M$ independent points $u^{(k)}$, $k=1,\ldots,M$, from a standard normal distribution.
\item Generate $M$ independent points $(\tilde{b}_i^{(k)},\tilde{b}_j^{(k)})$, as in SI section \ref{sect:noise}, which are left-tail associated
and which have standard-normal marginals. 
\item Again as in SI section \ref{sect:noise}, generate $M$ independent points $(\tilde{\ddot{b}}_i^{(k)},\tilde{\ddot{b}}_j^{(k)})$, 
$k=1,\ldots,M$, from a bivariate normal distribution with component variances $1$ and Pearson correlation the same as 
$(\tilde{b}_i^{(k)},\tilde{b}_j^{(k)})$.
\end{enumerate}
These data are stored and used repeatedly below. This provides computational efficiency because 
random number generation is computationally expensive.

Next, given values of the parameters $\delta$, $\mu_i - \mu_j$, and $\sigma$, proceed as follows to get estimates of the 
terms of the decomposition (\ref{eq:decomp_overall_IGR}).
\begin{enumerate}
\item Compute $\epsilon_i^0$ following (\ref{eq:decomp:baseline}), which is a simple formula of $\delta$ and 
$\mu_i - \mu_j$, not requiring use of the random samples described above. This gives an exact value of
$\epsilon_i^0$, not an estimate.
\item Estimate $\olep_i^E$, following (\ref{eq:decomp:contrib_of_var_E}), as 
$\widehat{\olep_i^E}=\mean_k [\ln(1-\delta+\delta \exp(\sigma u^{(k)}+\mu_i-\mu_j-\sigma^2/2))]-\epsilon_i^0$, an estimate with 
standard error $\se \left( \widehat{\olep_i^E} \right) = \sd_k [\ln(1-\delta+\delta \exp(\sigma u^{(k)}+\mu_i-\mu_j-\sigma^2/2))]/\sqrt{M}$,
where $\sd(\cdot)$ denotes sample standard deviation.
\item Estimate $\olep_i^C$, following (\ref{eq:decomp:contrib_of_var_C}), as 
$\widehat{\olep_i^C}=\mean_k [\ln(1-\delta+\delta \exp(-\sigma u^{(k)}+\mu_i-\mu_j+\sigma^2/2))]-\epsilon_i^0$, which has 
standard error $\se \left( \widehat{\olep_i^C} \right) = \sd_k[\ln(1-\delta+\delta \exp(-\sigma u^{(k)}+\mu_i-\mu_j+\sigma^2/2))]/\sqrt{M}$.
\item Note that $\tilde{b}_i^\shp-\tilde{b}_j^\shp$ in (\ref{eq:decomp:contrib_of_EC_var_per_se}) is normally distributed
with mean $0$ and variance $2$, so we can estimate $\olep_i^{(E \shp C)}$, following (\ref{eq:decomp:contrib_of_EC_var_per_se}),
as $\widehat{\olep_i^{(E \shp C)}}=\mean_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}+\mu_i-\mu_j))]-\widehat{\olep_i^E}-\widehat{\olep_i^C}-\epsilon_i^0.$ The standard error of this expression is 
$\se \left( \widehat{\olep_i^{(E \shp C)}} \right) = \sd_k [\ln(1-\delta+\delta\exp(\sigma \sqrt{2} u^{(k)}+\mu_i-\mu_j))-
\ln(1-\delta+\delta\exp(\sigma u^{(k)}+\mu_i-\mu_j-\sigma^2/2))-
\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\mu_i-\mu_j+\sigma^2/2))]/\sqrt{M}$. 
\item Estimate $\olep_i^{[EC]}$, following (\ref{eq:decomp:contrib_EC}), as
$\widehat{\olep_i^{[EC]}}=\mean_k[\ln(1-\delta+\delta\exp(\sigma(\tilde{\ddot{b}}_i^{(k)}-\tilde{\ddot{b}}_j^{(k)})+\mu_i-\mu_j))]-\mean_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}+\mu_i-\mu_j))]$. The standard error of this expression is 
$\se \left( \widehat{\olep_i^{[EC]}} \right) = \sqrt{s_1^2+s_2^2}$, where $s_1=\sd_k [\ln(1-\delta+\delta\exp(\sigma(\tilde{\ddot{b}}_i^{(k)}-\tilde{\ddot{b}}_j^{(k)})+\mu_i-\mu_j))]/\sqrt{M}$
and $s_2=\sd_k [\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}+\mu_i-\mu_j))]/\sqrt{M}$.
\item Estimate $\olep_i^{[E \nat C]}$, following (\ref{eq:decomp:contrib_ATA_2}), as 
$\widehat{\olep_i^{[E \nat C]}} = \mean_k[\ln(1-\delta+\delta\exp(\sigma(\tilde{b}_i^{(k)}-\tilde{b}_j^{(k)})+\mu_i-\mu_j))]-
\mean_k[\ln(1-\delta+\delta\exp(\sigma(\tilde{\ddot{b}}_i^{(k)}-\tilde{\ddot{b}}_j^{(k)})+\mu_i-\mu_j))].$ The standard
error of this expression is $\se \left( \widehat{\olep_i^{[E \nat C]}} \right)=\sqrt{s_3^2+s_1^2}$, where
$s_1$ is as above, and $s_3=\sd_k[\ln(1-\delta+\delta\exp(\tilde{b}_i^{(k)}-\tilde{b}_j^{(k)}))]/\sqrt{M}.$
\item Finally, estimate $\olr_{i \bs i}$, following (\ref{eq:base_IGR}), as 
$\widehat{\olr_{i \bs i}}=\mean_k[\ln(1-\delta+\delta\exp(\sigma(\tilde{b}_i^{(k)}-\tilde{b}_j^{(k)})+\mu_i-\mu_j))],$
which has standard error $s_3$ from above.
\end{enumerate}
It is easy to see that $\widehat{\olr_{i \bs i}}=\epsilon_i^0+\widehat{\olep_i^E}+\widehat{\olep_i^C}+\widehat{\olep_i^{(E \shp C)}}+
\widehat{\olep_i^{[EC]}}+\widehat{\olep_i^{[E \nat C]}},$ i.e., the decomposition (\ref{eq:decomp_overall_IGR}) applies 
at the level of the estimators, as well

Next, proceed as follows to get estimates of the terms in the decomposition (\ref{eq:decomp_overall_rjbsi}). 
\begin{enumerate}
\item Use $\epsilon_i^0=0$, as in (\ref{eq:decomp:baseline_j}), and this is the exact value.
\item Estimate $\olep_j^E$, following (\ref{eq:decomp:contrib_of_var_E_j}), as
$\widehat{\olep_j^E}=\mean_k[\ln(1-\delta+\delta\exp(\sigma u^{(k)}-\sigma^2/2))]-\epsilon_j^0$, an estimate
with standard error $\se \left( \widehat{\olep_j^E} \right) = \sd_k [\ln(1-\delta+\delta\exp(\sigma u^{(k)}-\sigma^2/2))]/\sqrt{M}$.
\item Estimate $\olep_j^C$, following (\ref{eq:decomp:contrib_of_var_C_j}), as
$\widehat{\olep_j^C}=\mean_k[\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]-\epsilon_j^0$, which has 
standard error $\se \left( \widehat{\olep_j^C} \right) = \sd_k [\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]/\sqrt{M}$.
\item Note that $\tilde{b}_j$ and $\tilde{b}_j^\shp$ in (\ref{eq:decomp:j:EshpC}) are standard normally distributed and
independent, so $\tilde{b}_j^\shp - \tilde{b}_j$ is normally distributed with mean $0$ and variance $2$. So we can 
estimate $\olep_j^{(E \shp C)}$, following (\ref{eq:decomp:j:EshpC}), as 
$\widehat{\olep_j^{(E \shp C)}} = \mean_k [\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))]-\widehat{\olep_j^E}-\widehat{\olep_j^C}-\epsilon_j^0$.
The standard error of this expression is $\se \left(  \widehat{\olep_j^{(E \shp C)}} \right) = 
\sd_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))-\ln(1-\delta+\delta\exp(u^{(k)}-\sigma^2/2))-
\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]/\sqrt{M}$.
\item Estimate
%***DAN: DAN, proceed from here
\end{enumerate}
